---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```




Let's do some light manipulation of the mtcars dataset, by summarising the variables.
This gives us an opportunity to explore the data and learn about pipes. 

First, let's take another look at the dataset using the `head()` function. Notice that there are only a few options for what the `carb` variable can be (1-8, although you can only see 1, 2, and 4 in these first 6 entries). It'd be useful to see
what the mean `mpg` is for each value of `carb`, so we can test whether these variables
are related. The `summarise()` command can be used here, which is part of the `dplyr`
package you loaded.  Type`?summarise` into the console to see the options for
how you can summarise the variables (e.g. mean, median, sd, IQR, min, max, etc).

For a simple count of how many cars there are, run

```{r summarise, exercise=TRUE}
summarise(mtcars, n())
```

The first argument in the function is the dataset, and the second is the summary function you'd like to use.
We can do more interesting things by grouping the data though. Let's count how many
cars are in each carb category by using the group_by() function:

```{r summarisegroup, exercise=TRUE}
summarise(group_by(mtcars,carb), carcount=n())
```

In this syntax, I have named the column we are creating with the number of cars `carcount`.
So there are 7 cars with `carb`=1, and 1 car with `carb`=8. Instead of just counting
the number of cars, we can explore the relationship between `carb` and `mpg` by calculating
the mean `mpg` of each of those groups of cars, and store this as the varible name `carstest`:

```{r carprogress, exercise=TRUE}
summarise(group_by(mtcars, carb), meanmpg=mean(mpg))
```

Alternatively, this same line of code can be written using the pipe operator, `%>%`.
You've seen this a few times, and can basically be thought of as the word 'goes into'.
So the code below is saying '`mtcars` goes into the function `group_by(XX, carb)` as the first
argument in the function'. That whole function then goes into the function
`summarise(XX, meanmpg=mean(mpg))` as the first argument.
This is exactly the same as the previous line of code, but some people find it easier
and neater to use the pipe operator, because you can then build the nested functions
more clearly. For now, use either form (but you will need to know how to read both).

```{r carprogresspipe, exercise=TRUE}
mtcars %>% # create a new object containing a summary of mtcars
  group_by(carb) %>% # group by the carbs variable (no. carburetors), and manufacturer
  summarise(meanmpg=mean(mpg)) # calculate mean mpg for each carb category
```

You can verify that carstest and cars are the same by typing each name into the console
to print them in their entirety.


```{r carprogressboth, exercise=TRUE}
cars_classic <- summarise(group_by(mtcars, carb), meanmpg=mean(mpg))

cars_tidy <- mtcars %>% # create a new object containing a summary of mtcars
  group_by(carb) %>% # group by the carbs variable (no. carburetors), and manufacturer
  summarise(meanmpg=mean(mpg)) # calculate mean mpg for each carb category

cars_classic
cars_tidy
```

You've stored these two datasets now as data frames with labelled columns, so you can
view each column separately using the `$`. In the embedded code chunk above, view the outputs from `cars_tidy$mpg` and `cars_tidy$carb`




