---
title: "Week 7:  Visualising High Dimensional Data    "
author: "Dr Kate Helmstedt"
date: "Sem 1, 2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
#library(kableExtra)
library(babynames)
library(viridis)
library(DT)
library(gridExtra)
library(ggrepel)
#library(fmsb)
#library(colormap) 
library(patchwork)
#library(GGally)
#library(ggquiver) 
library(gapminder)
library(ggalluvial)
library(emojifont)
library(palmerpenguins)
#library(hrbrthemes) #NOTE for mac users: if hrbrthemes is throwing an error for you, you need to install XQuartz https://www.xquartz.org/
```

# Setup

Load all the packages. Use `install.packages("")` to install any that don't load. I'd copy and run each line one by one
to see which you do and don't have the first time.


```{r remedy001}

library(tidyverse)
#library(kableExtra)
library(babynames)
library(viridis)
library(DT)
library(gridExtra)
library(ggrepel)
#library(fmsb)
#library(colormap) 
library(patchwork)
#library(GGally)
#library(ggquiver) 
library(gapminder)
library(ggalluvial)
library(emojifont)
library(palmerpenguins)
#library(hrbrthemes) #NOTE for mac users: if hrbrthemes is throwing an error for you, you need to install XQuartz https://www.xquartz.org/

```

# Some useful functions for manipulating data

Although this unit focuses on visulalising data, we often need to manipulate it too.

> Task: look up each of the following functions and add a short description of 
what they do and save it in an R markdown file (for your future reference only). 
Use `?` in the console, the help tab, rstudio cheat sheets, 
the R4ds book (https://r4ds.had.co.nz/) (we worked through this chapter in previous workshops) 
or the internet to look them up.

`%>%`
name: pipe
package: magrittr
function: 

`%in%`
name: match
package: base
function: 

`mutate()`
name:
package: dplyr
function: 

`filter()`
name:
package: dplyr
function: 

`select()`
name:
package: dplyr
function: 

`arrange()`
name:
package: dplyr
function: 

When you come across new functions in this workshop (e.g. `ifelse()`), add this to your list. Writing this list will help you keep things straight in your head throughout this workshop, and then might be a useful reference later.


# Faceting 
Data: `babynames`
Vis package: `ggplot2`

The following code creates a subset of the babynames dataset using `filter()`.

```{r remedy002}

US_names <- babynames::babynames %>% 
      filter(name %in% c("Mary","Emma", "Ida", "Ashley", "Amanda", "Jessica",    "Patricia", "Linda", "Deborah",   "Dorothy", "Betty", "Helen")) %>%
      filter(sex=="F")

```

Before we start playing with the data, take a quick look at it using the `head()` function to familiarise yourself with the variable names and data types.

## Spaghetti plot

```{r remedy003}

ggplot(US_names, aes(x=year, y=n, group=name, color=name)) +
      geom_line() +
      scale_color_viridis(discrete = TRUE) +
      theme(plot.title = element_text(size=14)) +
      ggtitle("A spaghetti chart of popular US names") 

```

## Highlighting 

It's a bit hard to see all the different lines, lets make it clearer and highlight one particular name, and label it.
Add a new variable (column) called `highlight` to the dataset using `mutate()`. 

```{r remedy004}

US_names_highlight <- US_names %>% 
      mutate(highlight = ifelse(name=="Amanda", "Amanda", "Other"))

```

Be patient! This one is a bit slow to load. 

```{r remedy005}

ggplot(US_names_highlight, aes(x=year, y=n, group=name, color=highlight, size=highlight)) +
      geom_line() +
      scale_color_manual(values = c("orange", "grey")) +
      scale_size_manual(values=c(1.5,0.2)) +
      theme(legend.position="none") +
      ggtitle("Popularity of US names in the previous 30 years") +
      geom_label( x=1990, y=55000, label="Amanda reached 3550 \n babies in 1970", size=4, color="orange") +
      theme(plot.title = element_text(size=14))

```

> Q. What does geom_label() do?
> Q. What does scale_size_manual() do?
> Task: go through each line of the code in the plot above. Do you know what each function is doing? 
If not, refer back to last week's workshop to find out. Copy this code to the document you're keeping with function names, and add a comment using `#` to each line about what
the code is doing. 


## Faceting 

Highlighting focuses on one line, but what if we want to see them all? Faceting can help with this. 

A faceted plot

```{r remedy006}

ggplot(US_names, aes(x=year, y=n, group=name)) +
  geom_line() +
  ggtitle("Popularity of US names in the previous 30 years") +
  facet_wrap(~name)

```

It's not the nicest plot, lets make some changes
> Task: Change geom_line() to geom_area()
> Task: add `fill = name` into aes() within ggplot()
> Task: change the colours using `scale_fill_brewer()` see the help file for palettes. 
> Task: Customise one other element of the plot using `theme()`. use the help menu to look up `theme()` 
and see the different arguments that can be used.

## Combine approaches 

Create a new dataset called `US_names_repeated`, make it the same as the `US_names` dataset but create one more 
variable (column) called `name_repeated`

```{r remedy006}

US_names_repeated <- US_names %>%
  mutate(name_repeated=name)

```

This was just to get around the fact that ggplot won't let us use the same variable for 
two things sometimes, so we just replicated a column so we can use it twice.

```{r remedy007}

ggplot(US_names_repeated, aes(x=year, y=n)) +
  
  # first, we will plot all the name curves in grey
  geom_line( data=US_names_repeated %>% dplyr::select(-name), aes(group=name_repeated), color="grey", size=0.5, alpha=0.5) +
  
  # then, we will plot only the name from that facet in blue
  geom_line( aes(color=name), color="69b3a2", size=1.2 )+
  
  # then, we add some visual elements to increase effectiveness
  scale_color_viridis(discrete = TRUE) +
  theme(
    plot.title = element_text(size=14),
    panel.grid = element_blank()
  ) +
  ggtitle("A spaghetti chart of popular US names") +
  
  # And then facet the data
  facet_wrap(~name)

```


# Multi-dimensional scatterplot (bubbleplot)

Data: `gapminder`
Vis package: `ggplot2`


Pepare the data: subset, order and add a new variable to the gapminder data using the code below.

```{r remedy009}

gapminder_data <- gapminder %>% 
  filter(year=="2007") %>% 
  dplyr::select(-year) %>%
  mutate(pop=pop/1000000) %>%
  arrange(desc(pop)) %>%
  mutate(country = factor(country, country))

```

> Task: go through each line of code above and make sure you understand what each line is doing to the data. 
> Q. why is `select()` written like this `dplyr::select()`?

Use `glimpse()` or `head()` to inspect and compare `gapminder` with `gapminder_data`.

```{r remedy010}

glimpse(gapminder)
glimpse(gapminder_data)
head(gapminder_data)

```

Generate bubble plot

```{r remedy011}

ggplot(gapminder_data, aes(x=gdpPercap, y=lifeExp, size = pop, color = continent)) +
    geom_point(alpha=0.7) +
    scale_size(range = c(1.4, 19), name="Population (M)") +
    scale_color_viridis(discrete=TRUE, guide=FALSE) +
    theme_bw() +
    theme(legend.position="none") 

```
  
> Task: Add a plot title and change the axis labels so they are clearer.
> Task: What could you remove from this code to make the country legend appear? What would you change to place the population bubble size legend at the bottom of the plot?
> Task: Remove the population variable from the plot. 
> Task: Create two different versions of this plot which map the variables differently. for example life expectancy on the x axis. 

Notice that we are using `theme_bw()` here, which is controlling the colours and design of the plot other than the data. Explore other options at https://ggplot2.tidyverse.org/reference/ggtheme.html, and play around with a few different looks.

## Emojis

Data: `palmerpenguins`
Vis package: `emojifont` and `ggplot2`

A simple dot plot without emojis

```{r remedy012}

ggplot(penguins, aes(bill_depth_mm, flipper_length_mm, color = species)) + 
  geom_point()

```

Re-create dot plot with emojis

```{r remedy013}

emoji_labels <- penguins %>%
  mutate(emojis = ifelse(species=='Adelie', emoji('bird'), ifelse(species=='Chinstrap', emoji('fish'), ifelse(species=='Gentoo', emoji('penguin'), NA))))

ggplot(emoji_labels, aes(bill_depth_mm, flipper_length_mm, color=species, label=emojis)) + 
      geom_text(family="EmojiOne", size=4)

```

Check out emoji choices here: https://apps.timwhitlock.info/emoji/tables/unicode


# Heat maps

Data: from Twitter using the `rtweet` package
Vis package: 
targeted_user <- "KateHelmstedt"
tweet_list <- get_timeline(targeted_user,n = 1200)
tweet_list <- tweet_list %>% 
    mutate(created_at = with_tz(created_at, tzone = "America/Los_Angeles"),
           week = week(created_at),
           weekday = wday(created_at, label = TRUE),
           year = year(created_at), 
           month = month(created_at, label = TRUE),
           date = date(created_at)) %>% 
    group_by(year, month) %>% 
    mutate(monthweek = week - min(week) + 1) %>% 
    group_by(date) %>% 
    mutate(activity = n(),
           original = if_else(is.na(reply_to_status_id)==TRUE &
                                  is_retweet==FALSE, TRUE, FALSE),
           likes = sum(favorite_count[original==TRUE]),
           retweets = sum(retweet_count[original==TRUE]))
calendarHeat(tweet_list$date, tweet_list$activity, 
             ncolors = 99, color = "g2r", varname="My Tweets & Replies Activitiy")



# Radar plot  
DATA: create dummy data
VIS PACKAGE: fms

Create synthetic data for the radar plot using the code below 
set.seed(1)
grades_data <- as.data.frame(matrix( sample( 2:20 , 10 , replace=T) , ncol=10))
colnames(grades_data) <- c("maths" , "english" , "biology" , "music" , "R-coding", "data-viz" , "french" , "physics", "statistics", "sport" )

Take a look at the format of the data we created
head(grades_data)

each subject in the dataset (column) has to have a max and min for the fms package to be able to draw the plot. the following line of code adds a row of 20 and a row of 0 to the data frame. The first argument `re(20,10)` add a row of 20 replicated 10 times.  

grades_data <- rbind(rep(20,10) , rep(0,10) , grades_data)

Create a basic radarChart
radarchart(grades_data, 
           title = "Spider Plot of Student Grades")

Add axis labels 
radarchart(grades_data, 
           title = "Spider Plot of Student Grades",
           axistype=1, specify the position of the label (select anumber between 0 and 6)
           caxislabels=seq(0,20,5) set the range of the axis labels
)

Change polygon and grid line colour
radarchart(grades_data, 
       title = "Spider Plot of Student Grades",
       axistype=1, axis label position (select anumber between 0 and 6)
       caxislabels=seq(0,20,5),set the range of the axis labels
       pcol = "forestgreen", polygon colour. 1 to 8, rgb code `rgb( 0.2,0.5,0.5,0.9)`, colour name
       cglcol="grey" gridline colour   
) 

There are many other functions that you can use to customise these plots. Here is an example. type `??radarplot` into the console to see a list of all the arguments. 
radarchart(grades_data, axistype=1, 
           pcol=rgb(0.2,0.5,0.5,0.9) ,  draw the polygon
           cglcol="grey", set the polygon colour
           cglty=1, define the line used to draw the polygon
           axislabcol="grey", define axis label colour
           caxislabels=seq(0,20,5), set the range of the axis labels
           cglwd=0.8, grid line width 
           vlcex=0.8, text size of labels 
           pfcol=rgb(0.2,0.5,0.5,0.5)
)

> Task: change the colour and thickness of the polygon line. 
> Task: Change one other feature (e.g grid line colour or thickness, or type of grid lind label. 
> Task: remove the shading inside the polygon.


Two polygons
-- How can we compare the grades of two students using a spider plot? One solution is to overlay them like the plot below. 

create data
grades_data_2 <-as.data.frame(matrix( c( sample( 2:20 , 10 , replace=T), sample( 2:9 , 10 , replace=T)) , ncol=10, byrow=TRUE))
colnames(grades_data_2) <- c("math" , "english" , "biology" , "music" , "R-coding", "data-viz" , "french" , "physic", "statistic", "sport" )
grades_data_2[2,2]=19

add min and max to each subject like the first spider plot
grades_data_2 <-rbind(rep(20,10) , rep(0,10) , grades_data_2)

Prepare polygon colours
colors_border = c(rgb(0.2,0.5,0.5,0.9), rgb(0.8,0.2,0.5,0.9))
colors_in = c(rgb(0.2,0.5,0.5,0.4), rgb(0.8,0.2,0.5,0.4))

Customise the radarChart
radarchart(grades_data_2, axistype=1,
            custom polygon
            pcol=colors_border , pfcol=colors_in , plwd=4, plty=1 , 
            custom the grid
            cglcol="grey", cglty=1, axislabcol="grey", caxislabels=seq(0,20,5), cglwd=1.1,
            custom labels
            vlcex=0.8)

add a Legend
legend(x=0.85, y=1, legend = c("Shirley", "Sonia"), bty = "n", pch=20 , col=colors_border , text.col = "black", cex=0.9, pt.cex=1.6)

But this won't work for more than 2 or 3 students. Facetting could address this for a larger dataset. 

PARALLEL COORDINATES PLOT  
DATA: iris
VIS PACKAGE: GGally (ggparcoord() function)

There is no geom in ggplot to create a parallel coordinate plots, so we will use ggparcoord from the GGally package. The structure of the arguments in ggparcoord() is designed to be very similar to ggplot 
-- Note: A arallel plot is the equivalent of a spider chart, but with cartesian coordinates. Thus, it is often prefered.

Lets have a look at the final plot to start with 
ggparcoord(iris, columns = 1:4, groupColumn = 5, order = "anyClass", 
           scale= "globalminmax", showPoints = TRUE, title = "Parallel Coordinate Plot for the Iris Data", alphaLines = 0.3) + 
    scale_color_viridis(discrete=TRUE) +
    theme(plot.title = element_text(size=10))

lets break this graph down to the minimum required elements 
ggparcoord(iris, columns = 1:4, groupColumn = 5) 

> Task: re-create the first graph by adding each argument or function one by one. Add a description of what each of them do below. 

arguments that go in aes()
1. order = :
2. showPoints = :
3. alphaLines = :
4. scale = :
functions
5. + scale_color_viridis() :
6. + theme(plot.title = element_text(size = )) :


> Task: Add x and y labs to the parrallel coordinates plot above. the xlab() and ylab() functions from ggplot will work here. 
> Task: The first plot uses `scale = "globalminmax". Use ??ggparcoord to find what other scales you can use. Try at least two different scales. 

NOTE: AXIS ORDER - optimizing the order of vertical axis can make the plot clearer. The general aim is to minimise the number of line crossings. 

See the next two figures, the second is much easier to interpret. Variable order is the only difference.

order = c(1:4)
ggparcoord(iris,
    columns = 1:4, groupColumn = 5, order = c(1:4),
    showPoints = TRUE, 
    title = "Original",
    alphaLines = 0.3
  ) + 
  scale_color_viridis(discrete=TRUE) +
  theme(
    legend.position="Default",
    plot.title = element_text(size=10)
  ) +
  xlab("")

order = "anyClass
ggparcoord(iris,
    columns = 1:4, groupColumn = 5, order = "anyClass",
    showPoints = TRUE, 
    title = "Re-ordered",
    alphaLines = 0.3
  ) + 
  scale_color_viridis(discrete=TRUE) +
  theme(
    legend.position="none",
    plot.title = element_text(size=10)
  ) +
  xlab("")

> Task: what other `order` can you use? See the help file for ggparcoord

ALLUVIAL DIAGRAM / SANKY PLOT 
DATA: UCBAdmissions
PACKAGE: ggalluvium()
NOTE: alluvial plots require some customisation, they are a little more tricky than the geoms in ggplot(). You will have seen some of the functions below, such as `scale_fill_brewer()` in last weeks workshop

UCB <- as.data.frame(UCBAdmissions)

have a quick look at the data
glimpse(UCB)
head(UCB)

One of the massive benefits of ggplot2 is that it can build plots up iteratively, by combining
multiple discrete functions using `+`. Throughout this workshop, we are going to take advantage of that
and assign parts of plots to variable names to minimise typing. Remember that in R, 
katerules<-10
assigns the number 10 to the variable name katerules. Now when I type
katerules
I just see 10. And I can use it as if it is 10:
katerules+3
Plots work the same. In this first example, we will assign a line plot to the name 'alluvial_0'.
Assigning it is not the same thing as running it! To then see the plot, we need to type the 
variable name into the console, which will print the saved plot.

map the variables using ggplot()
alluvial_0 <- ggplot(UCB, aes(y = Freq, axis1 = Gender, axis2 = Dept))
alluvial_0

add geom_alluvial(), use `fill = admit`` to add colour, and `width` to set the width of the layers (stratum)
alluvial_1 <- alluvial_0 + geom_alluvium(aes(fill = Admit), width = 1/12)
alluvial_1

add axis bars and label them. `width` sets the width of the vertical bars
alluvial_2 <- alluvial_1  + 
  geom_stratum(width = 1/12, fill = "black", color = "grey") + adds verticle bar with grey outline and black fill. 
  geom_label(stat = "stratum", infer.label = TRUE) labels the axis bars
alluvial_2

change labels on the x axis
alluvial_3 <- alluvial_2 +
  scale_x_discrete(limits = c("Gender", "Dept"), expand = c(.05, .05))
alluvial_3

Plot 4 - Change colour of layers 
alluvial_4 <- alluvial_3 + 
  scale_fill_brewer(type = "qual", palette = "Set1") 
alluvial_4

> Task: write out the entire final plot without any of the 'alluvial_' assigned variables, and verify it produces the same plot 
> Task: add a title to the plot. 
> Task: change the colour - use ??RColorBrewer to find a list of inbuilt palettes (half way down page)


VECTOR FIELDS 

Quiver plots of mathematical functions
A quiver plot displays velocity vectors as arrows with 
components (u,v) at the points (x,y).

First, explore some of the functions and elements we will use here:
(note that we aren't saving any of these, I'm just giving you an idea
of how we would explore what's happening in the code below -- useful
for this specific example, but also good practice every time you
come across a complex function with sub-components that you haven't seen
before)
seq(from=0,to=pi,length=13)
expand.grid(x=seq(from=0,to=10, length=11), 
            y=seq(from=0,to=30, length=11))

Now the actual function (remember that the pipe operator says
'use that as the first input variable in the next function)

First, we need to define all of the points at which we want an arrow to start 
-- this describes the density of the arrows in the final plot (you can play with this 
by changing the start point, end point, and the length of the vectors x and y here). 
There will be as many arrows as there are rows in the data, and each row of the data needs
to have two entries -- one giving the x position of the start of the arrow, and one
giving the y position of the start of the arrow.
expand.grid(x=seq(from=0,to=pi,length=13), 
            y=seq(from=0,to=pi,length=13)) %>%
  Feed that into a ggplot function that also defines the mathematical
  functions defining u and v, which determine the length and direction of the arrow
  ggplot(aes(x=x, y=y, u=cos(x), v=sin(y))) +
  And finally, feed these into the geom from the ggquiver package, geom_quiver
  geom_quiver()

As with any vector field, the functions for u and v can be any functional combination of x and y:
expand.grid(x=seq(from=0,to=10,length=11), 
            y=seq(from=0,to=10,length=11)) %>%
  ggplot(aes(x=x, y=y, u=x^2-y, v=x+y^2)) +
  geom_quiver()

TASK: play with different densities of arrows to generate plots of the vector field in the lecture slides on 
the slide titled "Main viz choice: density"
