---
title: 'MXB262 Workshop week 4: Effective Data Viz'
author: "Dr Kate Helmstedt"
date: 'Semester 1, 2021'
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
library(tidyverse)
#library(ggthemes)
library(viridis)
library(ggplot2)
#library(ggrepel)
library(ozmaps)
```

## Setup

The libraries we will use this week are:

```{r}
library(tidyverse)
library(ggthemes)
library(viridis)
library(ggplot2)
library(ggrepel)
library(ozmaps)

```

We need the names of the Australian states for some of what we're doing today, which can be found in the `ozmaps` package, so install and load that package too. The names of the states are contained in the object `ozmap_states$NAME`. State names for many different countries can be downloaded as data packages, so just have a google to see what you can find if you need different ones.

We will use the `covdat` package from last week too, so let's set that back up:

```{r, echo-TRUE, eval=TRUE}
  #remotes::install_github("kjhealy/covdata@main") # uncomment this if you need to install the package on a different computer this week
  library(covdata)

#  Australia's data
  mobility_australia <- apple_mobility %>% 
      filter(region == 'Australia', transportation_type == 'driving')
  
# Australia's data by state
  mobility_states <- apple_mobility %>% 
      filter(region %in% ozmap_states$NAME, transportation_type == 'driving') # note that I have used the `%in%` command here instead of the `==` above -- this   will filter for any words in the vector, which here is the whole list of Australian state names
```


## What is ` aes() ` 


 The `aes()` function describes how the data variables are mapped to the `geom_` function. 
 `aes()` can be set within `ggplot()`, which is how we have seen it so far. When placed here, the settings becomes the default 
 for all added `geom_()`. 
 Alternatively, it can be defined in the individual `geom_()` layers, and different settings can then apply to each `geom_`.
 For example the following two lines of code create the same plot 

### Including `aes()` within `ggplot()`

Create this plot...
```{r}
ggplot(mobility_australia,
           aes(date, score)) + 
  geom_line() +
  geom_point()

```

### Including aes() within `geom_()`

...and confirm the output is identical with this formulation:
```{r}
ggplot(mobility_australia) + 
  geom_line(aes(date, score)) +
  geom_point(aes(date, score)) # Note that we need to include the `aes()` here too since it isn't overridden in `ggplot()`
```

### Multiple lines through `aes()`

 If we have multiple datasets (e.g. multiple states), we can tell `ggplot` to 
 include them all through the `aes()` command.
 
```{r}
ggplot(mobility_states,
        aes(date, score, color=region)) +
  geom_line()  
```
 
 




## Manipulating plot features  


### Titles

Plot each of these to examine their effect:

#### Using ggtitle()

```{r}
ggplot(mobility_australia,
           aes(date, score)) + 
  geom_line()+ 
  ggtitle("People in Australia sure were chill in April") 

```

#### Titles over multiple lines
 
```{r}
ggplot(mobility_australia,
           aes(date, score)) + 
  geom_line()+ 
  ggtitle("People in Australia sure were chill \n in April") 
```

#### Add a subtitle
 
```{r}
ggplot(mobility_australia,
           aes(date, score)) + 
  geom_line()+ 
  ggtitle("People in Australia sure were chill in April", 
          subtitle = "Good work, Australia") 
```


#### Use lab() - an altentative to Titles and captions
 
```{r}
ggplot(mobility_australia,
           aes(date, score)) + 
  geom_line()+ 
  labs(title = "Boo, Covid-19 sucks", 
       subtitle = "But Australia is good at controlling it", 
       caption = "labs() lets me add a caption but ggtitle() doesn't")
```


#### Centre title
 
```{r}
ggplot(mobility_australia,
           aes(date, score)) + 
  geom_line()+ 
  ggtitle("In April I was sad :(") +
  theme(plot.title = element_text(hjust = 0.5 ))
```


#### Adjust text size, make bold, add a margin around title
 
```{r}
ggplot(mobility_australia,
           aes(date, score)) + 
  geom_line()+ 
  ggtitle("In April many of us were sad :(")  + 
  theme(plot.title = element_text(hjust = 0.5, size = 20, 
                                  face = "bold", margin = margin(10,0,10,0) ))
```


 > NOTE: you can also add titles, subtitles etc within ` theme() `. 
 You can see more on themes here - https://www.datanovia.com/en/blog/ggplot-title-subtitle-and-caption/
 
 
### Legends

 Remember that `<-` is how we assign (or store) values or snippets of code to a 
 variable name. You can do this with plots too! Remember though that the line that 
 assigns or stores will NOT then also run the code or print the result -- you need 
 to them type the variable name in order to see the output

For example

 First store it
 
`test1 <- 3 + 4`

 Then display it
 
`test1`

 Then use it
 
`test1 + 5`

 First store it
```{r}
g <- ggplot(mobility_states,
        aes(date, score, color=region)) +
  geom_line()  
```
 

 Then display it
 
`g`

 Then use it:

#### Turn off legend title

```{r}
g + theme(legend.title = element_blank()) 
```



#### Change legend colour, size and to bold

```{r}
g + theme(legend.title = element_text(colour="chocolate", size=16, face="bold")) 
```

#### Change legend title

Here, we need to match the input (i.e. color) to the
 attribute we assigned it to above. See how we used `color = country` here? We
 now use `labs(color = )` to change the title of the legend.
 
```{r}
g + labs(color = "This is a different legend title") 
```
 
#### Everything together 

```{r}
ggplot(mobility_states,
        aes(date, score, color=region)) +
  geom_line()   + 
  theme(legend.title = element_text(colour="chocolate", size=16, face="bold")) + 
  labs(color = "Australian State")
```

### Axes

#### Axis labels 

```{r}
ggplot(mobility_states,
        aes(date, score, color=region)) +
  geom_line() + 
  labs(x = "Time", y = "Cumulative confirmed cases") 

```

 > NOTE: I'll create an object called `axis_labels` so I don't have to type so much.
 Not only am I kind of lazy, but this also helps to make more readable code. The person 
 reading the code just has to read once how I created axis_labels, understand and agree
 with it once, and then doesn't have to see all that code every time.

 First store it

```{r}
axis_labels <- ggplot(mobility_states,
        aes(date, score, color=region)) +
  geom_line()  + 
  labs(x = "Time", y = "Cumulative confirmed cases") 
```

 Then display it

```{r}
 
axis_labels

```

 Then use it:

#### Customise angle shape and size using `theme()` 

```{r}
axis_labels + theme(axis.text.x = element_text(angle = 50, size = 10, vjust = 0.5))
```

#### Change colour of axis title and labels

```{r}
 
axis_labels + theme(
      axis.text.x = element_text(angle = 50, size = 10, vjust = 0.5, colour = 'forestgreen'), 
      axis.title.x = element_text(colour = "blue"))

```

### Maniulate axes

#### Truncate axes 
#######################################################

```{r}
 
ggplot(mobility_australia,
           aes(date, score)) + 
  geom_line() + xlim(as.date("2020-01-13", "2020-06-08"))

```

 Flip Axes

```{r remedy007}
 
ggplot(top_global_counts, aes(y=confirm, x=name)) + 
geom_bar(position="dodge", stat="identity") 

```

 flipped

```{r remedy008}
 
ggplot(top_global_counts, aes(y=confirm, x=name)) + 
  geom_bar(position="dodge", stat="identity") +
  coord_flip()

```

 -- TASK: Use the method from last week to sort this bar plot 

 Scatterplot options
 to avoid extra typing, create an object called dots and assign the ggplot() function 
 to it so that you just have to add the different geom layers. 
dots <- ggplot(Australia, aes(time, cum_confirm)) 

 colour 

```{r remedy009}
 
dots + geom_point(color="firebrick")

```

 transparency 

```{r remedy010}
 
dots + geom_point(alpha = 0.2)

```

 size 

```{r remedy011}
 
dots + geom_point(size = 5)

```

 shape 
 
```{r remedy012}

dots + geom_point(shape = 4)  # add a shape name or a number for example 'cross' or '4'. 


```
 --TASK:how many different shapes can you find?

 All together 

```{r remedy013}
 
dots + geom_point(colour = "firebrick", size = 3, alpha = 0.5, shape = 5)

```

 Background

 Colour 

```{r remedy014}
 
ggplot(Australia,aes(time, cum_confirm)) +
  geom_line()   + theme(panel.background = element_rect(fill = 'grey75')) 

```

 --TASK: Uncomment the previous plot to change the background colour to a darker grey.   

 Grid lines 

```{r remedy015}
 
ggplot(Australia,aes(time, cum_confirm)) +
  geom_line() + 
  theme(panel.grid.major = element_line(colour = "yellow", size=2),
        panel.grid.minor = element_line(colour = "blue"))

```

 Margins

```{r remedy016}
 
ggplot(Australia,aes(time, cum_confirm)) +
  geom_line() + 
  theme(plot.background=element_rect(fill="darkseagreen"))

```

 all together

```{r remedy017}
 
ggplot(Australia,aes(time, cum_confirm)) +
  geom_line() + 
  theme(panel.grid.major = element_line(colour = "yellow", size=2),
        panel.grid.minor = element_line(colour = "purple"), 
        plot.background = element_rect(fill = "darkseagreen"))


```


  Themes  


 `theme()` is a useful layer that can be used to customise features of a plot. 
 You can customize the non-data aspects of the plot with themes. 
 Such as font type, font size, title position, subtitle colour, add a caption and many others
 This is a way to keep all plots within your
 document, website, thesis, or publication the same. This of it as a coded style guide -- plots will keep the
 theme and be recognisable as parts of a whole.

```{r remedy018}

ggplot(Australia,aes(time, cum_confirm)) +
  geom_line() + 
  theme_bw()

```

 ggplot2 has 8 built in themes
 TASK: re-plot the graph with another built-in theme (type "theme_" and options will pop up)

 You can also control the individual components of the plot within a theme, like text size:

```{r remedy019}

ggplot(Australia) +
  geom_line(aes(time, cum_confirm)) + 
  theme(text = element_text(size = 16))

```

 There's a bunch more in the ggthemes() package created by Jeffrey Arnold. Install and load the package.

 The theme from fivethirtyeight.com

```{r remedy020}
 
ggplot(Australia) +
  geom_line(aes(time, cum_confirm)) + 
  theme_fivethirtyeight()

```

 The theme from The Economist

```{r remedy021}
 
ggplot(Australia) +
  geom_line(aes(time, cum_confirm)) + 
  theme_economist()


```
 The theme from Excel (why?)

```{r remedy022}
 
ggplot(Australia) +
  geom_line(aes(time, cum_confirm)) + 
  theme_excel()

```

 -- TASK: re-create the plot above using a different theme. Type `theme_` and select tab, a drop down list of themes will appear. 


 Colours 
library(RColorBrewer)
 Colour must be mapped to a variable using aes(), this can be in ggplot() or in geom(). 
 After you've identified which variable will be coloured you can then use one of the 
 `scale......()` functions to define the colours. 

 map colour using aes() within ggplot() - outline colour

```{r remedy023}
 
colour_outline <- ggplot(iris, aes(Species, Sepal.Length, colour = Species)) +
  geom_boxplot()

```
  
colour_outline  note that I created `colour_outline` so we don't have to keep typing the ggplot() function over and over again. 


 add colour using aes() within geom() - fill colour

```{r remedy024}
 
colour_fill <- ggplot(iris, aes(Species, Sepal.Length)) + 
  geom_boxplot(aes(fill = Species))  
colour_fill

```

 --NOTE: Colour has not been specified in the previous two plots. The colours in the plots are default settings. 


```{r remedy025}
 specify colour by adding  ` + scale_colour_manual` (outline) or `+ scale_fill_manual` (fill) 
colour_outline + scale_colour_manual(values = c("00AFBB", "E7B800", "FC4E07"))
colour_fill + scale_fill_manual(values = c("00AFBB", "E7B800", "FC4E07"))

```

 specify colour using a palette. Here we use an existing palette. But you can also create your own. 

```{r remedy026}
 
colour_fill + scale_fill_brewer(palette = "Dark2")
colour_outline + scale_colour_brewer(palette = "Dark2")  

```

see other palettes here - https://www.r-graph-gallery.com/38-rcolorbrewers-palettes/ 

```{r remedy027}

ggplot(ToothGrowth, aes(x = supp, y = len)) + geom_point() + geom_jitter(width = 0.2) + scale_colour_gradient()
 

```
 Colours available in R - http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf

 Vertical and Horizontal Lines 

```{r remedy028}

ggplot(Australia) +
  geom_line(aes(time, cum_confirm))  +
  geom_hline(yintercept = 1000)

```

 Annotations: It can be useful to label individual observations, and groups.
 Replace the legend with in-text

```{r remedy029}
 
irislabels <- iris %>%
  group_by(Species) %>%
  summarize_all(mean)

ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, color = Species)) + 
  geom_point() + geom_text(data  = irislabels, aes(x=Sepal.Length, y=Sepal.Width, label = Species)) + 
  theme(legend.position = "none")


```
 Add a figure number

```{r remedy030}
 
ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, color = Species)) + 
  geom_point() + annotate("text", x=4.2, y=4.4, label = "A")

```

 TASK: Make the annotated "A" larger



 
    Overplotting       

 We're going to create a dataset for this section of the workshop -- it 
 intentionally has overplotting problems, but you'll also see these kinds 
 of problems in your real-world and simulation datasets.

 Dataset:

```{r remedy031}
 
a <- data.frame( x=rnorm(20000, 10, 1.2), y=rnorm(20000, 10, 1.2), group=rep("A",20000))
b <- data.frame( x=rnorm(20000, 14.5, 1.2), y=rnorm(20000, 14.5, 1.2), group=rep("B",20000))
c <- data.frame( x=rnorm(20000, 9.5, 1.5), y=rnorm(20000, 15.5, 1.5), group=rep("C",20000))
synthetic_data <- do.call(rbind, list(a,b,c))               

```

 Oveplotting is common and occurs when your dataset is too large. 
 See the following scatterplot of the data we just created.  

```{r remedy032}
 
ggplot(synthetic_data, aes(x=x, y=y)) +
  geom_point() 

```

 There doesn't seem to be a relationship between x and y. But if you 
 explore the following steps for addressing overplotting, you will 
 see a relationship emerge. 

 First, when we suspect there might be an overplotting issue, we can view the 
 point density using the ggplotpointdensity package. Install it and load the
 library now. Then:

```{r remedy033}
 
ggplot(synthetic_data, aes(x=x, y=y)) +
  geom_pointdensity() 

```

 We can see areas of high density here (in lighter blue with this default
 colour scheme). Now that we have identified an overplotting issue, there are a 
 couple of options for viewing the data in a more representitive way:

 1. Dot size

```{r remedy034}
 
ggplot(synthetic_data, aes(x=x, y=y)) +
  geom_point(size = 0.2) 

```

 2. Transparency

```{r remedy035}
 
ggplot(synthetic_data, aes(x=x, y=y)) +
  geom_point(size = 0.2, alpha = 0.01) 

```

 -- Q. what argument manipulated transparency? is it in ggplot() or in geom_point()?
 -- Q. increase the dot size, does this improve the plot?
 -- TASK: try a few different combinations of `alpha` and `size`.

 3. Sampling
 Sometimes less is more. Plotting a subset of the data (5% here) can help avoid overplotting, 
 and can also greatly reduce the computing time.   

 take a sample of the data

```{r remedy036}
 
synthetic_data_subset <- sample_frac(synthetic_data, 0.05)

ggplot(synthetic_data_subset, aes(x=x, y=y)) +
  geom_point(size=2)

```

 -- Q. did this help you see the data? Did you lose much information? When might this not be a good idea?

 4. Grouping with colour

```{r remedy037}
 
ggplot(synthetic_data, aes(x=x, y=y, color=group)) +
  geom_point( size=2, alpha=0.1) +
  scale_color_viridis(discrete=TRUE) 

```

 -- NOTE: the colour argument is within the ggplot( aes() ) function and not in geo_point(). 
 Also that `group` is a variable in the data set (look back up at where we simulated it), 
 and not a specific colour. 

 5. Highlighting just one group 
 This assumes there is a grouping variable in the data set. 

```{r remedy038}
 
ggplot(synthetic_data, aes(x=x, y=y)) +
  geom_point(color="grey", size=2) +
  geom_point(data = synthetic_data %>% filter(group=="B"), color="69b3a2", size=2)

```

 -- Q. what argument in geom_point() changes the colour? Test what other common colour names work. 
 -- Q does the Australian/UK spelling of colour work as well?
 -- Q can you put the colour argument in the ggplot() function instead of geom_point()?
 -- Q. Can the colour argument in geom_point() be set to `group`, similarly to No. 4 above?

 6. Jittering 
 Here's the example of jittering that we saw last week. This is another way to deal with overplotting

```{r remedy039}

ggplot(ToothGrowth, aes(x = supp, y = len)) + geom_point() 
ggplot(ToothGrowth, aes(x = supp, y = len)) + geom_point() + geom_jitter(width = 0.2)

```






